---
title: "Proposal"
author: "Ahmed Hassan, Joel Martin, Joseph Strobel, Rabsa Naseer, Wil Hillmer, Sierra Hoff"
date: "10/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Date Source
[NCAA Men's Basketball Dataset](https://www.kaggle.com/andrewsundberg/college-basketball-dataset?select=cbb.csv)

### Overview
```{r}
df <- read.csv('datasets/cbb20.csv')
head(df)
```


This data set includes team stats for all Division 1 Men's basketball teams for the years 2015-2020. Each year resides in its respective csv file. Some of the more interesting stats for basketball include points, rebounds, assists, turnovers and shooting percentage which this dataset includes. Additionally, one interesting column this dataset includes is 'Wins Above Bubble.' This will be interesting to view as a reference point when determining which stats are indicitive of making the NCAA March Madness Touranament.

### Plan for Exploratory Analysis
Since this dataset includes all Division One teams, we will look at which stats correspond with higher achieving teams. We can do this for all teams and then specifically focus in on Iowa State to see how they perform over the years. It will be very interesting to see over the time period how each team performs and to determine why or why they aren't making it to the end of season tournament. 
